---
title: "Measles and Rubella Cases"
format: 
  dashboard:
    orientation: columns
    theme: sandstone
    logo: who-logo.png
    logo-alt: "World Health Organization logo with a wreath and a snake. "
    nav-buttons:
      - text: Nicole's LinkedIn
        icon: linkedin
        href: https://www.linkedin.com/in/nicole-yee-203279313/
        aria-label: LinkedIn
      - text: Hailey's LinkedIn
        icon: linkedin
        href: https://www.linkedin.com/in/hailey-ernest/ 
        aria-label: LinkedIn
embed-resources: true
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(gt)
library(xfun)
library(httr)
library(tidyjson)
library(jsonlite)
library(rvest)
library(chromote)
library(hexbin)
library(ggridges)
library(viridis)
library(patchwork)
library(glue)
library(stringr)
library(stringi)
library(rsconnect)
```

```{r}
#| label: load-data
#| include: false

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv') |>
  clean_names()

cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv') |>
  clean_names()

gdp_web <- readr::read_csv('gdp_web.csv')
```

```{r}
#| label: prep-gdp-data
#| include: false

clean_country <- function(x) {
  x |>
    stri_trans_general("Latin-ASCII") |>
    str_to_lower() |>
    str_replace_all("&", "and") |>
    str_remove_all("[[:punct:]]") |>
    str_squish()
}

gdp_web_clean <- gdp_web |>
  rename(year = scraped_year) |>
  mutate(
    year = as.numeric(year),
    country_join = recode(country,
      "United States" = "United States of America",
      "Bolivia" = "Bolivia (Plurinational State of)",
      "Venezuela" = "Venezuela (Bolivarian Republic of)"
    ),
    country_key = clean_country(country_join)
  )

cases_year_clean <- cases_year |>
  mutate(
    year = as.numeric(year),
    country_key = clean_country(country)
  )

cases_year_gdp <- cases_year_clean |>
  left_join(
    gdp_web_clean |>
      select(country_key, year, gdp_per_capita),
    by = c("country_key", "year")
  )

cases_gdp <- cases_month |>
  left_join(
    cases_year_gdp,
    by = c("country", "year"),
    suffix = c("_month", "_year")
  ) |>
  as_tibble()

cases_gdp_year <- cases_year |>
  filter(year != 2025) |>
  mutate(
    year = as.numeric(year),
    measles_epi_linked = as.numeric(measles_epi_linked)
  ) |>
  left_join(
    gdp_web |>
      rename(year = scraped_year) |>
      select(country, year, gdp_per_capita),
    by = c("country", "year")
  ) |>
  mutate(
    region = recode(region,
      "AFRO" = "Africa",
      "AMRO" = "Americas",
      "EMRO" = "Eastern Mediterranean",
      "EURO" = "Europe",
      "SEARO" = "South-East Asia",
      "WPRO" = "Western Pacific"
    )
  ) |>
  filter(
    !is.na(measles_epi_linked),
    !is.na(gdp_per_capita),
    measles_epi_linked > 0
  )
```

## Column {width=48%}

::: {.callout-note appearance="minimal"}
**Research Question 1**

Globally, has there been improvement in containing measles and rubella cases in recent years, and is there a time of year where these cases become more common?
:::

### Research Question 1: Measles and Rubella Cases Over Time

::: {.panel-tabset}

#### Yearly trends

```{r}
#| label: yearly-trends
#| warning: false
#| message: false
#| fig-height: 6
#| fig-width: 11

annotate_text <- tibble(
  year = c(2014.5, 2014.3),
  y_lab = c(75000, 200000),
  disease = c("Rubella", "Measles")
)

cases_year |>
  group_by(year) |>
  summarize(
    measles = sum(measles_total, na.rm = TRUE),
    rubella = sum(rubella_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot() +
  geom_ribbon(
    aes(x = year, ymin = measles, ymax = rubella),
    position = "identity",
    inherit.aes = FALSE,
    fill = "#F8F5F0"
  ) +
  geom_line(aes(x = year, y = measles, color = "Measles"), linewidth = 1) +
  geom_line(aes(x = year, y = rubella, color = "Rubella"), linewidth = 1) +
  scale_color_manual(
    name = "Disease",
    values = c("Measles" = "#325D88", "Rubella" = "#3E3F3A")
  ) +
  scale_x_continuous(breaks = unique(cases_year$year), expand = c(0.01, 0.01)) +
  scale_y_continuous(
    breaks = seq(0, 600000, by = 100000),
    labels = scales::label_comma()
  ) +
  geom_text(
    data = annotate_text,
    mapping = aes(x = year, y = y_lab, label = disease, color = disease),
    inherit.aes = FALSE,
    show.legend = FALSE
  ) +
  theme_bw() +
  labs(
    x = "Year",
    y = "Number of Cases",
    title = "Measles and Rubella Cases Over Time",
    subtitle = "Similar spikes occur across diseases, but measles cases are more volatile."
  ) +
  theme(panel.grid.minor = element_blank())
```

#### Monthly patterns

```{r}
#| label: monthly-patterns
#| warning: false
#| message: false
#| fig-height: 7
#| fig-width: 12

color_gradient <- colorRampPalette(c("#A6DAF5", "#325D88"))(12)
names(color_gradient) <- month.name

measles <- cases_month |>
  group_by(month, year) |>
  summarise(measles_total = sum(measles_total, na.rm = TRUE), .groups = "drop") |>
  mutate(month_full = month.name[month]) |>
  ggplot(aes(
    x = measles_total,
    y = reorder(month_full, -month),
    group = month,
    fill = month_full
  )) +
  geom_density_ridges() +
  scale_x_continuous(breaks = seq(0, 140000, by = 40000), labels = scales::label_comma()) +
  scale_fill_manual(values = color_gradient) +
  scale_y_discrete(expand = c(0.0001, 0.5)) +
  theme_minimal() +
  labs(x = "Number of Cases", y = "", title = "Measles") +
  theme(panel.grid.minor.x = element_blank(), legend.position = "none")

color_gradient <- colorRampPalette(c("#B8B9B6", "#3E3F3A"))(12)
names(color_gradient) <- month.name

rubella <- cases_month |>
  group_by(month, year) |>
  summarise(rubella_total = sum(rubella_total, na.rm = TRUE), .groups = "drop") |>
  mutate(month_full = month.name[month]) |>
  ggplot(aes(
    x = rubella_total,
    y = reorder(month_full, -month),
    group = month,
    fill = month_full
  )) +
  geom_density_ridges() +
  scale_x_continuous(breaks = seq(0, 20000, by = 4000), labels = scales::label_comma()) +
  scale_fill_manual(values = color_gradient) +
  scale_y_discrete(expand = c(0.0001, 0.5)) +
  theme_minimal() +
  labs(x = "Number of Cases", y = "", title = "Rubella") +
  theme(panel.grid.minor.x = element_blank(), legend.position = "none")

(measles + rubella) +
  plot_layout(guides = "collect", axis_titles = "collect") +
  plot_annotation(
    title = "Comparison of Global Measles and Rubella Cases for Each Month",
    subtitle = "Monthly distributions span case counts from 2012 to 2025"
  )
```

#### Peak months table

```{r}
#| label: peak-months-table
#| warning: false
#| message: false

get_max_month <- function(yr) {
  cases_month |>
    filter(year == yr) |>
    group_by(month) |>
    summarize(
      num_measles = sum(measles_total, na.rm = TRUE),
      num_rubella = sum(rubella_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    summarize(
      measles_month = month[which.max(num_measles)],
      measles_cases = max(num_measles),
      rubella_month = month[which.max(num_rubella)],
      rubella_cases = max(num_rubella)
    ) |>
    mutate(year = yr, .before = 1)
}

cases_year |>
  distinct(year) |>
  pull(year) |>
  map_dfr(get_max_month) |>
  mutate(
    measles_month = month.name[measles_month],
    rubella_month = month.name[rubella_month]
  ) |>
  gt() |>
  cols_label(
    year = md("**Year**"),
    measles_month = md("**Month**"),
    measles_cases = md("**Cases**"),
    rubella_month = md("**Month**"),
    rubella_cases = md("**Cases**")
  ) |>
  tab_header(title = md("Time of Year with Increased Measles and Rubella Cases Over Time")) |>
  tab_caption(caption = md("Months with the highest number of cases.")) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_spanner(label = "Measles", columns = c(measles_month, measles_cases)) |>
  tab_spanner(label = "Rubella", columns = c(rubella_month, rubella_cases)) |>
  fmt_number(columns = c(3, 5), sep_mark = ",", decimals = 0)
```

:::

## Column {width=52%}

::: {.callout-note appearance="minimal"}
**Research Question 2**

Are countries with lower GDP per capita associated with higher numbers of epidemiologically linked measles cases, and does this relationship vary by region?
:::

### Research Question 2: Measles Association with GDP per Capita

::: {.panel-tabset}

#### GDP vs. cases

```{r}
#| label: gdp-vs-cases
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 13

ggplot(cases_gdp_year,
       aes(x = year, y = measles_epi_linked, color = gdp_per_capita)) +
  geom_jitter(alpha = 0.9, size = 2.5, width = 0.25, height = 0.1) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = c(2012, 2016, 2020, 2024)) +
  scale_color_gradient(low = "#A6DAF5", high = "#3E3F3A", labels = scales::comma) +
  facet_wrap(~ region, ncol = 2, scales = "free_y") +
  labs(
    title = "Epidemiologically Linked Measles Cases Over Time by Region",
    subtitle = "Each point represents a country in a year; color shows GDP per capita",
    x = "Year",
    y = "Epidemiologically linked measles cases (log scale)",
    color = "GDP per capita"
  ) +
  theme_minimal(base_size = 16)
```

#### Epi-linked proportions

```{r}
#| label: epi-proportion-table
#| warning: false
#| message: false

# function to get proportion of epidemiologically-linked cases for each year in one region
epi_proportions <- function(region_code) {
  cases_month |>
    filter(.data$region == region_code) |>
    group_by(year) |>
    summarise(
      measles_epi_linked = sum(measles_epi_linked, na.rm = TRUE),
      measles_lab_confirmed = sum(measles_lab_confirmed, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      prop_epi_measles = measles_epi_linked / 
        (measles_epi_linked + measles_lab_confirmed)
    ) |>
    select(year, prop_epi_measles)
}

cases_month |>
  distinct(region) |>
  mutate(
    region_name = recode(region,
      "AFR" = "Africa",
      "AMR" = "Americas",
      "EMR" = "Eastern Mediterranean",
      "EUR" = "Europe",
      "SEAR" = "South-East Asia",
      "WPR" = "Western Pacific"
    )
  ) |>
  mutate(epi_props = map(region, epi_proportions)) |>
  unnest(epi_props) |>
  select(region_name, year, prop_epi_measles) |>
  group_by(region_name) |>
  gt() |>
  fmt_percent(columns = matches("prop"), decimals = 1) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_row_groups()) |>
  cols_label(
    region_name = md("**Region**"),
    year = md("**Year**"),
    prop_epi_measles = md("**Epidemiologically-linked measles cases**")
  ) |>
  tab_header(title = md("Proportion of Confirmed Cases that are Epidemiologically-linked")) |>
  tab_caption(caption = md("Changes in each region's confirmed measles cases makeup over time."))
```


#### GDP groups

```{r}
#| label: gdp-group-table
#| warning: false
#| message: false

cases_gdp_year_global <- cases_gdp_year |>
  mutate(
    gdp_group = ntile(gdp_per_capita, 3),
    gdp_group = recode(as.character(gdp_group),
      "1" = "Low GDP",
      "2" = "Medium GDP",
      "3" = "High GDP"
    ),
    gdp_group = factor(gdp_group, levels = c("Low GDP", "Medium GDP", "High GDP"))
  )

gdp_epi_groups <- function(region_name) {
  cases_gdp_year_global |>
    filter(.data$region == region_name) |>
    group_by(gdp_group) |>
    summarise(epi_linked_cases = sum(measles_epi_linked, na.rm = TRUE), .groups = "drop")
}

cases_gdp_year |>
  distinct(region) |>
  mutate(gdp_epi = map(region, gdp_epi_groups)) |>
  unnest(gdp_epi) |>
  complete(region, gdp_group, fill = list(epi_linked_cases = 0)) |>
  group_by(region) |>
  gt() |>
  fmt_number(columns = epi_linked_cases, decimals = 0) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_row_groups()) |>
  cols_label(
    region = md("**Region**"),
    gdp_group = md("**GDP Level**"),
    epi_linked_cases = md("**Epidemiologically-linked measles cases**")
  ) |>
  tab_header(title = md("Epidemiologically-linked Measles Cases by Regional GDP Level")) |>
  tab_caption(caption = md("Total epidemiologically-linked measles cases grouped by global GDP per capita level."))
```

#### Regional distributions

```{r}
#| label: regional-distributions
#| warning: false
#| message: false
#| fig-height: 8
#| fig-width: 13

epi_region_plot <- cases_gdp_year |>
  ggplot(aes(x = log1p(measles_epi_linked), y = region, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_x_continuous(
    breaks = log1p(c(0, 10, 100, 1000, 10000)),
    labels = scales::comma(c(0, 10, 100, 1000, 10000))
  ) +
  scale_fill_gradient(low = "#F8F5F0", high = "#325D88") +
  scale_y_discrete(expand = c(0.0001, 0.5)) +
  labs(x = "Epidemiologically linked measles cases (log scale)", y = "", fill = "Cases") +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor.x = element_blank(), legend.position = "none")

gdp_region <- cases_gdp_year |>
  ggplot(aes(x = gdp_per_capita, y = region, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_fill_gradient(low = "#A6DAF5", high = "#3E3F3A") +
  scale_y_discrete(expand = c(0.0001, 0.5)) +
  labs(x = "GDP per capita (log scale)", y = "", fill = "GDP per capita") +
  theme_minimal(base_size = 16) +
  theme(panel.grid.minor.x = element_blank(), legend.position = "none")

epi_region_plot + gdp_region +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Distributions of Measles Cases and GDP per Capita by Region",
    subtitle = "Yearly country distributions from 2012-2024"
  )
```

#### Model results

```{r}
#| label: model-results
#| warning: false
#| message: false

epi_gdp_model_data <- cases_gdp |>
  distinct(country, year, region_year, gdp_per_capita, measles_epi_linked_year) |>
  rename(region = region_year) |>
  filter(year != 2025, gdp_per_capita > 0) |>
  drop_na(gdp_per_capita, measles_epi_linked_year, region) |>
  mutate(
    region = factor(region),
    log_gdp = log10(gdp_per_capita),
    log_epi_linked = log1p(measles_epi_linked_year)
  )

epi_gdp_model <- lm(log_epi_linked ~ log_gdp * region, data = epi_gdp_model_data)

coef_table <- as.data.frame(summary(epi_gdp_model)$coefficients) |>
  rownames_to_column(var = "term")

coef_table |>
  gt() |>
  fmt_number(columns = c(Estimate, `Std. Error`, `t value`, `Pr(>|t|)`), decimals = 3) |>
  sub_small_vals(columns = `Pr(>|t|)`, threshold = 0.001) |>
  cols_align(align = "center") |>
  cols_label(
    term = "Model Term",
    Estimate = "Estimate",
    `Std. Error` = "Standard Error",
    `t value` = "t Statistic",
    `Pr(>|t|)` = "p-value"
  ) |>
  tab_header(
    title = "Model Results: GDP, Region, and Epidemiologically Linked Measles Cases",
    subtitle = "Log-log linear regression model with GDP by region interaction"
  ) |>
  tab_caption(
    caption = "Relationship between GDP per capita and epidemiologically linked measles cases using yearly country observations from 2012-2024."
  ) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())
```

:::
