---
title: "Measles and Rubella Cases"
format: 
  dashboard:
    orientation: rows
    theme: sandstone
    logo: who-logo.png
    logo-alt: "World Health Organization logo with a wreath and a snake. "
    nav-buttons:
      - text: Nicole's LinkedIn
        icon: linkedin
        href: https://www.linkedin.com/in/nicole-yee-203279313/
        aria-label: LinkedIn
      - text: Hailey's LinkedIn
        icon: linkedin
        href: https://www.linkedin.com/in/hailey-ernest/ 
        aria-label: LinkedIn
server: shiny
embed-resources: true
---

```{r}
#| label: load-packages
#| include: false
#| context: setup

library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(gt)
library(xfun)
library(httr)
library(tidyjson)
library(jsonlite)
library(rvest)
library(chromote)
library(hexbin)
library(ggridges)
library(viridis)
library(patchwork)
library(glue)
library(stringr)
library(stringi)
library(rsconnect)
library(leaflet)
library(plotly)
library(shiny)
library(DT)
library(htmltools)
```

```{r}
#| label: load-data
#| include: false
#| context: setup

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv') |>
  clean_names()

cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv') |>
  clean_names()

gdp_web <- readr::read_csv('gdp_web.csv')

lat_lon <- read_csv(here::here("world-data-2023.csv")) |>
  mutate(Country = recode_values(Country,
                                 "Republic of the Congo" ~ "Congo",
                                 "The Gambia" ~ "Gambia",
                                 "The Bahamas" ~ "Bahamas",
                                 "Czech Republic" ~ "Czechia",
                                 "Republic of Ireland" ~ "Ireland",
                                 default = Country))
```

```{r}
#| label: prep-gdp-data
#| include: false
#| context: setup

clean_country <- function(x) {
  x |>
    stri_trans_general("Latin-ASCII") |>
    str_to_lower() |>
    str_replace_all("&", "and") |>
    str_remove_all("[[:punct:]]") |>
    str_squish()
}

gdp_web_clean <- gdp_web |>
  rename(year = scraped_year) |>
  mutate(
    year = as.numeric(year),
    country_join = recode(country,
      "United States" = "United States of America",
      "Bolivia" = "Bolivia (Plurinational State of)",
      "Venezuela" = "Venezuela (Bolivarian Republic of)"
    ),
    country_key = clean_country(country_join)
  )

cases_year_clean <- cases_year |>
  mutate(
    year = as.numeric(year),
    country_key = clean_country(country)
  )

cases_year_gdp <- cases_year_clean |>
  left_join(
    gdp_web_clean |>
      select(country_key, year, gdp_per_capita),
    by = c("country_key", "year")
  )

cases_gdp <- cases_month |>
  left_join(
    cases_year_gdp,
    by = c("country", "year"),
    suffix = c("_month", "_year")
  ) |>
  as_tibble()

cases_gdp_year <- cases_year |>
  filter(year != 2025) |>
  mutate(
    year = as.numeric(year),
    measles_epi_linked = as.numeric(measles_epi_linked)
  ) |>
  left_join(
    gdp_web |>
      rename(year = scraped_year) |>
      select(country, year, gdp_per_capita),
    by = c("country", "year")
  ) |>
  mutate(
    region = recode(region,
      "AFRO" = "Africa",
      "AMRO" = "Americas",
      "EMRO" = "Eastern Mediterranean",
      "EURO" = "Europe",
      "SEARO" = "South-East Asia",
      "WPRO" = "Western Pacific"
    )
  ) |>
  filter(
    !is.na(measles_epi_linked),
    !is.na(gdp_per_capita),
    measles_epi_linked > 0
  )|>
  left_join(lat_lon, by = join_by("country" == "Country"))
```

# Measles by GDP Worldwide

::: {.callout-note appearance="minimal" style="background-color: #F0F8FF; border-left: 6px solid #0093D5;"}
**Research Question**

Are countries with lower GDP per capita associated with higher numbers of epidemiologically linked measles cases, and does this relationship vary by region?
:::

### Research Question 2: Measles Association with GDP per Capita

:::::: panel-tabset

#### Countries with High GDP and Measles Cases

```{r}
#| layout: [[20,80]]
#| layout-valign: top

selectInput(
  inputId = "gdp_year",
  label = "Select a year below to view the countries with the highest GDP per capita or lowest measles cases per 1,000,000 people in the population:",
  choices = 2013:2024,
  selected = 2024,
  width = "100%"
)

leafletOutput(
  "map",
  height = "480px",
  width = "100%"
)
```


```{r}
#| label: high-gdp-measles-map
#| warning: false
#| message: false
#| context: server

top_measles <- cases_gdp_year |>
  group_by(year) |>
  slice_min(n=20,
            order_by = measles_incidence_rate_per_1000000_total_population) |>
  mutate(type = "cases")

 output$map <- renderLeaflet({
    cases_gdp_year |>
  group_by(year) |>
  slice_max(n=20, order_by = gdp_per_capita) |>
  mutate(type = "gdp") |>
  bind_rows(top_measles) |>
  filter(year == input$gdp_year) |>
  group_by(country, Longitude, Latitude) |>
  summarise(both = n() == 2,
            type = if_else(both, "both", first(type)),
            gdp = first(gdp_per_capita),
            measles_cases = first(measles_incidence_rate_per_1000000_total_population),
            .groups = "drop") |>
  mutate(circle_label = 
           glue::glue(
             "{country}<br>
             GDP per capita: {scales::comma(gdp)}<br>
             Measles incidence rate: {measles_cases}")) |>
  leaflet(
    options = leafletOptions(
      minZoom = 2,
      maxBoundsViscosity = 1
    )
  ) |>
    addTiles(
      options = tileOptions(noWrap = TRUE)
    ) |>
    setView(
      lng = 0,
      lat = 20,
      zoom = 2
    ) |>
    setMaxBounds(
      lng1 = -180,
      lat1 = -60,
      lng2 = 180,
      lat2 = 85
    ) |>
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    label = ~lapply(circle_label, htmltools::HTML),
    color = ~case_when(
      type == "both" ~ "#8B3A3A",
      type == "gdp" ~ "#325D88",
      type == "cases" ~ "#968777"
    ),
    radius = 6,
    stroke = TRUE,
    fillOpacity = 1
    # clusterOptions = markerClusterOptions
  ) |>
  addLegend(
    position = "bottomleft",
    colors = c("#325D88", "#968777", "#8B3A3A"),
    labels = c("Highest GDP per capita", 
               "Lowest Measles Cases per 1,000,000 population",
               "Both in highest GDP per capita and lowest cases per 1,000,000 people"),
    title = "Top Countries (2024)"
  )
   })

```

::: {.callout-note appearance="minimal" style="border-left: 4px solid #325D88;"}
Displayed above are the 20 countries with highest GDP per capita and the 20 countries with the lowest number of measles cases per 1,000,000 people in it's population for one year. There is a notable amount of overlap in each year. For example, **twelve countries belong to both groups** in 2024. Over time, most countries in these groups belong to **Europe or Africa**.
:::

#### GDP per capita levels

```{r}
#| label: gdp-group-table
#| warning: false
#| message: false
#| fig-width: 16
#| fig-height: 16

cases_gdp_year_global <- cases_gdp_year |>
  mutate(
    gdp_group = ntile(gdp_per_capita, 5),
    gdp_group = recode(as.character(gdp_group),
      "1" = "Low GDP",
      "2" = "Lower-Middle GDP",
      "3" = "Middle GDP",
      "4" = "Upper-Middle GDP",
      "5" = "High GDP"
    ),
    gdp_group = factor(
      gdp_group,
      levels = c(
        "Low GDP",
        "Lower-Middle GDP",
        "Middle GDP",
        "Upper-Middle GDP",
        "High GDP"
      )
    )
  )

gdp_incidence_table <- cases_gdp_year_global |>
  group_by(region, gdp_group) |>
  summarise(
    median_incidence_rate = median(
      measles_incidence_rate_per_1000000_total_population,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  complete(
    region,
    gdp_group,
    fill = list(median_incidence_rate = 0)
  ) |>
  group_by(region) |>
  mutate(
    region_rank = percent_rank(median_incidence_rate)
  ) |>
  ungroup() |>
  select(region, gdp_group, median_incidence_rate) |>
  pivot_wider(
    names_from = gdp_group,
    values_from = median_incidence_rate
  )

gdp_incidence_table |>
  gt() |>
  fmt_number(
    columns = c(
      `Low GDP`,
      `Lower-Middle GDP`,
      `Middle GDP`,
      `Upper-Middle GDP`,
      `High GDP`
    ),
    decimals = 2
  ) |>
  data_color(
  columns = c(
    `Low GDP`,
    `Lower-Middle GDP`,
    `Middle GDP`,
    `Upper-Middle GDP`,
    `High GDP`
  ),
    direction = "row",
    palette = c("#F8F5F0", "#C8BBA3", "#7A6F63")
  ) |>
  cols_label(
    region = md("**Region**"),
    `Low GDP` = md("**Low**"),
    `Lower-Middle GDP` = md("**Lower-Middle**"),
    `Middle GDP` = md("**Middle**"),
    `Upper-Middle GDP` = md("**Upper-Middle**"),
    `High GDP` = md("**High**")
  ) |>
  tab_spanner(
    label = md("**GDP Group**"),
    columns = c(
      `Low GDP`,
      `Lower-Middle GDP`,
      `Middle GDP`,
      `Upper-Middle GDP`,
      `High GDP`
    )
  ) |>
  tab_header(
    title = md("Measles Incidence Rates by Regional GDP Level")
  ) |>
  tab_caption(
    caption = md("Median incidence rates per 1,000,000 population grouped by GDP per capita quintiles, with each GDP group representing approximately 20% of observations. Table colors compare values within each region.")
  ) |>
  tab_options(
    table.width = pct(55),
    table.font.size = px(16),
    data_row.padding = px(12),
    heading.padding = px(12) 
  ) |>
  cols_align(
    align = "center",
    columns = c(
      `Low GDP`,
      `Lower-Middle GDP`,
      `Middle GDP`,
      `Upper-Middle GDP`,
      `High GDP`
    )
  )
```

::: {.callout-note appearance="minimal"}
Median measles incidence rates vary across regions and GDP groups. The Americas show consistently **low incidence rates** across all GDP groups, and Europe has the **highest incidence rates among lower-middle GDP countries**. Africa and South-East Asia show **high incidence rates in upper-middle GDP groups**, and the Western Pacific has its highest incidence rates among **lower-middle GDP countries**.
:::

#### GDP vs. Epidemiologically linked cases

```{r}
#| label: gdp-vs-cases
#| warning: false
#| message: false
#| fig-height: 12
#| fig-width: 22

ggplot(cases_gdp_year,
       aes(x = year, y = measles_epi_linked, color = gdp_per_capita)) +
  geom_jitter(alpha = 0.9, size = 4, width = 0.25, height = 0.1) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = c(2012, 2016, 2020, 2024)) +
  scale_color_gradient(low = "#A6DAF5", high = "#3E3F3A", labels = scales::comma) +
  facet_wrap(~ region, ncol = 3, scales = "free_y") +
  labs(
    title = "Epidemiologically Linked Measles Cases Over Time by Region",
    subtitle = "Each point represents a country in a year, color shows GDP per capita, 
    and case counts shown on a log scale",
    x = "Year",
    y = NULL,
    color = "GDP per capita"
  ) +
  theme_minimal(base_size = 30) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1.2, "lines"),
    plot.margin = margin(4, 4, 4, 4)
  )
```

::: {.callout-note appearance="minimal"}
Africa and South-East Asia show **wide ranges** of epidemiologically linked measles cases and are generally associated with lower GDP countries. The Americas report relatively few linked cases across years, while Europe shows varying outbreak levels with little clear GDP relationship. In contrast, the Eastern Mediterranean and Western Pacific regions display a stronger trend where **lower GDP countries** tend to experience **higher measles case counts**.
:::
::::::

# Measles in Ethiopia

::: {.callout-note appearance="minimal" style="background-color: #F0F8FF; border-left: 6px solid #0093D5;"}
*Why Ethiopia?* Ethiopia was selected because it showed the **largest** increase (about 25,000 more yearly cases) in epidemiologically linked measles cases from 2012 to 2024. Ethiopia’s measles incidence rate **increased** by about 181 cases per 1,000,000 people over the study period, which was one of the highest in Africa. This case study focuses on Ethiopia’s outbreak pattern over time and uses average temperature as additional information for exploring possible environmental patterns.

Data on Ethiopia’s yearly temperature was obtained through the [World Bank Group’s Climate Change Knowledge Portal](https://climateknowledgeportal.worldbank.org/country/ethiopia/climate-data-historical). The [data](https://crudata.uea.ac.uk/cru/data/hrg/) was collected by the Climatic Research Unit of University of East Anglia from ground-based weather station observations and can be accessed.
:::

## 

::::: panel-tabset
#### Epidemiologically linked case increase

```{r}
#| label: ethiopia-outlier
#| warning: false
#| message: false
#| fig-width: 19
#| fig-height: 10

outlier_country <- "Ethiopia"

plot_data <- cases_gdp_year |>
  filter(region == "Africa")

highlight_data <- plot_data |>
  filter(country == outlier_country)

ggplot(
  plot_data,
  aes(x = year, y = measles_epi_linked, color = gdp_per_capita)
) +
  geom_jitter(
    alpha = 0.45,
    size = 4.5,
    width = 0.25,
    height = 0.05
  ) +

  geom_point(
    data = highlight_data,
    aes(x = year, y = measles_epi_linked),
    inherit.aes = FALSE,
    color = "#968777",
    fill = "#968777",
    size = 8,
    shape = 21,
    stroke = 1.2
  ) +

  scale_y_log10(
    limits = c(1, 100000),
    labels = scales::comma
  ) +

  scale_x_continuous(
    breaks = c(2012, 2016, 2020, 2024)
  ) +

  scale_color_gradient(
    low = "#A6DAF5",
    high = "#3E3F3A",
    labels = scales::comma
  ) +

  labs(
    title = "Ethiopia's Notable Increase in Measles Cases Over Time",
    subtitle = "Ethiopia has one of the largest country's increase in epidemiologically linked measles cases over time.\nCase counts shown on a log scale",
    x = "Year",
    y = NULL,
    color = "GDP per capita ($)"
  ) +

  theme_minimal(base_size = 27) +

  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

::: {.callout-note appearance="minimal"}
Ethiopia shows an **increase** in epidemiologically linked measles cases over time from 2012 to 2024. Interestingly, from 2012 to 2015 cases were rising, but there is a big decrease from 2016 to 2021 before a drastic increase by 2024.
:::

```{r}
#| include: false
cases_gdp_year |>
  group_by(country) |>
  summarize(
    first_cases = first(measles_epi_linked),
    last_cases = last(measles_epi_linked),
    change_cases = last_cases - first_cases,
    .groups = "drop"
  ) |>
  arrange(desc(change_cases)) |>
  slice_head(n = 15)
```

```{r}
#| include: false
cases_gdp_year |>
  arrange(country, year) |>
  group_by(country, region) |>
  summarise(
    first_year = first(year),
    last_year = last(year),
    first_incidence = first(measles_incidence_rate_per_1000000_total_population),
    last_incidence = last(measles_incidence_rate_per_1000000_total_population),
    incidence_change = last_incidence - first_incidence,
    .groups = "drop"
  ) |>
  arrange(desc(incidence_change)) |>
  slice_head(n = 15)
```

#### Temperature vs. Epi-linked cases

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 8

library(plotly)

ethiopia_temp <- readr::read_csv("ethiopia_temp.csv") |>
  clean_names() |>
  rename(
    year = 1,
    avg_temp_c = 2
  ) |>
  mutate(year = as.numeric(year))

ethiopia_cases <- cases_gdp_year |>
  filter(country == "Ethiopia", year >= 2012, year <= 2024) |>
  select(year, measles_epi_linked)

ethiopia_plot_data <- ethiopia_cases |>
  left_join(ethiopia_temp, by = "year") |>
  filter(!is.na(avg_temp_c))

plot_ly(
  ethiopia_plot_data,
  x = ~year,
  y = ~measles_epi_linked,
  type = "bar",
  marker = list(
    color = ~avg_temp_c,
    colorscale = list(
      c(0, "#A6DAF5"),
      c(0.5, "#F4D06F"),
      c(1, "#8B3A3A")
    ),
    colorbar = list(title = "Avg temp °C")
  ),
  hovertemplate = paste(
    "<b>Ethiopia</b><br>",
    "Year: %{x}<br>",
    "Epi-linked measles cases: %{y:,}<br>",
    "Avg temperature: %{marker.color:.2f}°C",
    "<extra></extra>"
  )
) |>
  layout(
    title = "Ethiopia's Epidemiologically Linked Measles Cases and Temperature Over Time",
    xaxis = list(title = "Year", dtick = 1),
    yaxis = list(title = "")
  )
```

::: {.callout-note appearance="minimal"}
Ethiopia’s epidemiologically linked measles cases were high in 2014 and 2015, and increased sharply after 2021, reaching the highest level in 2024. The temperature coloring shows that **warmer** years tend to align with the outbreak pattern, thus **rising measles counts** tend to have much warmer yearly temperature conditions.
:::
:::::

# Data

```{r}
#| height: 30%
#| layout: [[50, 50]]
#| layout-valign: top

htmltools::div(
  style = "background-color: #EEF3F6;
  padding: 10px 14px;
  border-radius: 10px;
  height: 125px;",

  htmltools::tags$div(
    style = "display: flex; align-items: center; gap: 10px; margin-bottom: 2px;",

    htmltools::img(
      src = "who-logo.png",
      height = "34px"
    ),

    htmltools::h4(
      "WHO Measles & Rubella Data",
      style = "margin: 0;"
    )
  ),

  htmltools::p(
    style = "margin-top: 2px; margin-bottom: 0px; line-height: 1.2; font-size: 16px;",
    htmltools::HTML(
      'Monthly and yearly summaries of suspected and confirmed measles and rubella cases across countries and regions from 2012 to 2024. Data are collected through national surveillance systems and compiled by WHO. Cases are classified as clinically compatible, epidemiologically linked, or laboratory-confirmed. Source: <a href="https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-06-24/readme.md" target="_blank">TidyTuesday GitHub</a>.'
    )
  )
)

htmltools::div(
  style = "background-color: #F9F6F1;
  padding: 10px 14px;
  border-radius: 10px;
  height: 125px;",

  htmltools::tags$div(
    style = "display: flex; align-items: center; gap: 10px; margin-bottom: 2px;",

    htmltools::span(
      htmltools::HTML('<i class="bi bi-cash-stack"></i>'),
      style = "font-size: 26px;",
      color = "#5C4033"
    ),

    htmltools::h4(
      "World Bank GDP Data",
      style = "margin: 0;"
    )
  ),

  htmltools::p(
    style = "margin-top: 2px; margin-bottom: 0px; line-height: 1.2; font-size: 16px;",
    htmltools::HTML(
      'GDP per capita data for over 200 countries from 1990 to 2024. GDP per capita measures economic output relative to population and is commonly used to compare economic development across countries. Source: <a href="https://ourworldindata.org/grapher/gdp-per-capita-worldbank?tab=table&time=1991..latest&tableFilter=countries" target="_blank">Our World in Data</a>.'
    )
  )
)
```

```{r}
#| height: 70%
#| warning: false
#| message: false

cases_gdp_year |>
  select(
    country,
    region,
    year,
    gdp_per_capita,
    measles_total,
    measles_epi_linked,
    measles_incidence_rate_per_1000000_total_population
  ) |>
  DT::datatable(
    rownames = FALSE,
    filter = "top",
    options = list(
      pageLength = 10,
      scrollX = TRUE,
      scrollY = "430px"
    )
  )
```